Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-12 paper/paper.qmd
Minor revisions to proof read text

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities. Projected coral
    cover were analysed for each reefs ability to maintain positive carbonate production budgets under
    future conditions using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

@@

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While Grose et al. [-@grose2023] do not
consider oceanic data, they provide a region-specific selection mechanism which is used as
an indicator of general performance of GCMs for the region. An overview of the selected GCMs
is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed GFDL-ESM4,
whereas Dixon et al., [-@dixon2022; -@dixon2023] provides data based on GFDL-CM4. This
substitution is assumed appropriate given they share the same fundamental representation of
the physical climate system. The use of these GCMs therefore allows for a range of possible
futures to be explored, capturing uncertainty in future responses to CO~2~ increases
(@fig-methods-dhws; @sfig-ecs;[@grose2023; @shepherd2018]) and enables the assessment of
modelled coral reef behaviour over a broad range of independently derived climate futures
[@abramowitz2019; @brunner2020].

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model [GBR-1 version 2.0; @steven2019], as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

@@

Comparing across GCMs, it is common for reefs to switch between clusters, including from Low
to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2197 reefs, 21% (462) of
reefs have consistent cluster assignment across all GCMs. Of the 462 consistently assigned
reefs, 23% are within the Far Northern Management area, with 1%, 55% and 21% occurring in
Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management areas respectively
(@sfig-gcm-cluster-map). Of the 462 consistently assigned reefs, 34%, 18% and 47% are
clustered into Low, Medium and High clusters respectively. High levels of assignment
variability across GCMs are expected as due to spatial differences in each given GCM’s DHW
trajectories.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by management area scale cluster assignment.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Reefs that do not have a consistent cluster
assignment level across all GCMs at mangement area scale are coloured grey. All consistently
clustered reefs are coloured according to their management area scale cluster assignment.
Green represents reefs in the low cover cluster, orange represents reefs in the medium cover
cluster and blue represents reefs in the high cover cluster. All 2197 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/gcm_cluster_assignment_map.png){#sfig-gcm-cluster-map
fig-align="center" width="14.82cm"}

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

@@

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While Grose et al. [-@grose2023] do not
consider oceanic data, they provide a region-specific selection mechanism which is used as
an indicator of general performance of GCMs for the region. An overview of the selected GCMs
is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed GFDL-ESM4,
whereas Dixon et al., [-@dixon2022; -@dixon2023] provide data based on GFDL-CM4. This
substitution is assumed appropriate given they share the same fundamental representation of
the physical climate system. The use of these GCMs therefore allows for a range of possible
futures to be explored, capturing uncertainty in future responses to CO~2~ increases
(@fig-methods-dhws; @sfig-ecs; @grose2023; @shepherd2018) and enables the assessment of
modelled coral reef behaviour over a broad range of independently derived climate futures
[@abramowitz2019; @brunner2020].

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model (GBR-1 version 2.0; @steven2019), as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

@@

Comparing across GCMs, it is common for reefs to switch between clusters, including from Low
to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2197 reefs, 21% (462) of
reefs have consistent cluster assignment across all GCMs. Of the 462 consistently assigned
reefs, 23% are within the Far Northern Management area, with 1%, 55% and 21% occurring in
Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management areas respectively
(@sfig-gcm-cluster-map). Of the 462 consistently assigned reefs, 34%, 18% and 47% are
clustered into Low, Medium and High clusters respectively. High levels of assignment
variability across GCMs are expected due to spatial differences in each given GCM’s DHW
trajectories.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by management area scale cluster assignment.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Reefs that do not have a consistent cluster
assignment level across all GCMs at management area scale are coloured grey. All
consistently clustered reefs are coloured according to their management area scale cluster
assignment. Green represents reefs in the low cover cluster, orange represents reefs in the
medium cover cluster and blue represents reefs in the high cover cluster. All 2197 reefs
included in this investigation are represented in this figure. Annotations represent major
coastal settlements along the
GBR.](../figs/gcm_cluster_assignment_map.png){#sfig-gcm-cluster-map fig-align="center"
width="14.82cm"}
